Speed and Precision in Range Analysis

### 2012-04-17 doc/Manuscript/SBLP/paper.tex
Fixed erros in the SBLP paper.

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is partially ordered by the subset relation
$\sqsubseteq$.
$\varset{Z}^2$ is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
The objective of range analysis is to determine a mapping
$I: \varset{V} \mapsto \varset{Z}^2$ from the set of integer program variables
$V$ to intervals, such that, for any variable $v \in V$, if
$I(v) = [l, u]$, the, during the execution of the target program, any
valued $i$ assigned to $v$ is such that $l \leq i \leq u$.

@@

\noindent
\textbf{Growth analysis: }
The objective of growth analysis is to determine the growth behavior of
each program variable.
There are four possible behaviors:
(a) the variable is bound to a constant interval,
such as $k_0$ in Figure~\ref{fig:overall_view}(b).
(b) The variable is bound to a decreasing interval, i.e., an interval whose
lower bound decreases.
This is the case of $j_1$ in our example.
(c) The variable is bound to an increasing interval, i.e., its upper bound
increases.
This is the case of $i_1$ in the example.
(d) The variable is bound to an interval that expands in both directions.
The growth analysis uses an infinite lattice, i.e., $\varset{Z}^2$.
Thus, a careless implementation of an algorithm that infers growth patterns
might not terminate.
In order to ensure termination, we must reckon on a technique called
{\em widening}, first introduced by Cousot and Cousot as a key component
of abstract interpretation~\cite{Cousot77}.
There are many different widening strategies.
We discuss some of them in Section~\ref{sub:widen}.

\noindent
\textbf{Future resolution: }
In order to learn information from comparisons between variables, such as
\texttt{i < j} in Figure~\ref{fig:overall_view}(a), we bind some intervals
to {\em futures}.
Futures are symbolic limits, which will be replaced by actual numbers once
we finish the growth analysis.
The ranges found by the growth analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the growth analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$ be analyzed before $V$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:overall_view}(b), variables $j_1$ and $i_t$
are in the same SCC only because of the control dependence edges.

@@

\noindent
\textbf{Narrowing Analysis.}
The growth analysis associates very conservative bounds to each variable.
Thus, the last step of our algorithm consists in narrowing these intervals.
We accomplish this step via Cousot and Cousot's classic narrowing
operator~\cite[248]{Cousot77}, which we show in
Figure~\ref{fig:crop_analysis}.

@@

\noindent
\textbf{Example:}
Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strong component of
Figure~\ref{fig:overall_view}(d).
Upon meeting this SCC, we have already determined that the interval
$[0, 0]$ is bound to $i_0$ and that the interval $[100, 100]$ is bound to
$j_0$.
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:overall_view}(d).
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $k_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, by only analyzing $j$'s SCC after we have
analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.

@@

The experiments also reveal that the memory consumption of our implementation
is linear with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities together.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the heaviest benchmarks, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\textwidth]{images/MemCorr}
\end{center}
\caption{\label{fig:MemCorr}
Comparison between program size (number of var nodes in constraint
graphs after inlining) and memory consumption (KB).
Coefficient of determination = 0.9947.
}
\end{figure}

@@

As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
However, whereas that approach have not been used with programs larger than
the Stanford benchmark suite, we, as shown before, have been able to
deal with remarkably large programs.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/impactSCC}
\end{center}
\caption{\label{fig:impactSCC}
(Left) Bars give time to run our analysis without building strong components
divided by time to run the analysis on strongly connected components.
(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
components, dividid by the precision that we obtain without them.
}
\end{figure}

@@

A naive implementation of range analysis would be intra-procedural; that is,
would solve the range analysis problem once for each function.
However, we can gain in precision by performing it inter-procedurally.
An inter-procedural implementation allows the results found for a function $f$
to flow into other functions that $f$ calls.
Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
program seen in Figure~\ref{fig:overall_view}(a).
The trivial way to produce an inter-procedural implementation is to insert
into the constraint system assignments from the actual parameter names to the
actual parameter names.
In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
of information from $0$, the actual parameter, to $k_0$, the formal parameter
of function \texttt{foo}.

@@

Another way to increase the precision of range analysis is by using a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
A usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worst precision results.
On the other hand, for the SPEC programs inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/wholeImpact}
\end{center}
\caption{\label{fig:wholeImpact}
The impact of whole program analysis on precision. Each bar gives precision in
\%bitwidth reduciton.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
\end{center}
\caption{\label{fig:timeComp}
Runtime comparison between the intra, inter e inter+inline versions of
our algorithm.
The bars are normalized to the time to run the intra-procedural analysis.
}
\end{figure}

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation are responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our runtime, because it increases the
size of the target program.
We have tried our analysis in two different program representations: the
Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
Single Assignment (e-SSA) form~\cite{Bodik00}.
The SSA form gives us a faster, albeit more imprecise, analysis.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/wideningPrec}
\end{center}
\caption{\label{fig:wideningPrec}
The impact of the widening strategy on precision.
Each bar gives precision in \%bitwidth reduciton.}
\end{figure}

>>>>

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is partially ordered by the subset relation
$\sqsubseteq$.
$\varset{Z}^2$ is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
The objective of range analysis is to determine a mapping
$I: \varset{V} \mapsto \varset{Z}^2$ from the set of integer program variables
$V$ to intervals, such that, for any variable $v \in V$, if
$I(v) = [l, u]$, then, during the execution of the target program, any
valued $i$ assigned to $v$ is such that $l \leq i \leq u$.

@@

\noindent
\textbf{Growth analysis. }
The objective of growth analysis is to determine the growth behavior of
each program variable.
There are four possible behaviors:
(a) the variable is bound to a constant interval,
such as $k_0$ in Figure~\ref{fig:overall_view}(b).
(b) The variable is bound to a decreasing interval, i.e., an interval whose
lower bound decreases.
This is the case of $j_1$ in our example.
(c) The variable is bound to an increasing interval, i.e., its upper bound
increases.
This is the case of $i_1$ in the example.
(d) The variable is bound to an interval that expands in both directions.
The growth analysis uses an infinite lattice, i.e., $\varset{Z}^2$.
Thus, a careless implementation of an algorithm that infers growth patterns
might not terminate.
In order to ensure termination, we must reckon on a technique called
{\em widening}, first introduced by Cousot and Cousot as a key component
of abstract interpretation~\cite{Cousot77}.
There are many different widening strategies.
We discuss some of them in Section~\ref{sub:widen}.

\noindent
\textbf{Future resolution. }
In order to learn information from comparisons between variables, such as
\texttt{i < j} in Figure~\ref{fig:overall_view}(a), we bind some intervals
to {\em futures}.
Futures are symbolic limits, which will be replaced by actual numbers once
we finish the growth analysis.
The ranges found by the growth analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the growth analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$ be analyzed before $V$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:overall_view}(b), variables $j_1$ and $i_t$
are in the same SCC only because of the control dependence edges.

@@

\noindent
\textbf{Narrowing analysis.}
The growth analysis associates very conservative bounds to each variable.
Thus, the last step of our algorithm consists in narrowing these intervals.
We accomplish this step via Cousot and Cousot's classic narrowing
operator~\cite[248]{Cousot77}, which we show in
Figure~\ref{fig:crop_analysis}.

@@

\noindent
\textbf{Example.}
Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strong component of
Figure~\ref{fig:overall_view}(d).
Upon meeting this SCC, we have already determined that the interval
$[0, 0]$ is bound to $i_0$ and that the interval $[100, 100]$ is bound to
$j_0$.
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:overall_view}(d).
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $k_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, by only analyzing $j$'s SCC after we have
analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.

@@

The experiments also reveal that the memory consumption of our implementation
is linear with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities together.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\textwidth]{images/MemCorr}
\end{center}
\caption{\label{fig:MemCorr}
Comparison between program size (number of var nodes in constraint
graphs) and memory consumption (KB).
Coefficient of determination = 0.9947.
}
\end{figure}

@@

As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
However, whereas that approach have not been used with programs larger than
the Stanford benchmark suite, we, as shown before, have been able to
deal with remarkably larger programs.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/impactSCC}
\end{center}
\caption{\label{fig:impactSCC}
(Left) Bars give time to run our analysis without building strong components
divided by time to run the analysis on strongly connected components.
(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
components, divided by the precision that we obtain without them.
}
\end{figure}

@@

A naive implementation of range analysis would be intra-procedural; that is,
would solve the range analysis problem once for each function.
However, we can gain in precision by performing it inter-procedurally.
An inter-procedural implementation allows the results found for a function $f$
to flow into other functions that $f$ calls.
Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
program seen in Figure~\ref{fig:overall_view}(a).
The trivial way to produce an inter-procedural implementation is to insert
into the constraint system assignments from the actual parameter names to the
formal parameter names.
In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
of information from $0$, the actual parameter, to $k_0$, the formal parameter
of function \texttt{foo}.

@@

Another way to increase the precision of range analysis is by using a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worst precision results.
On the other hand, for the SPEC programs, inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/wholeImpact}
\end{center}
\caption{\label{fig:wholeImpact}
The impact of whole program analysis on precision. Each bar gives precision in
\%bitwidth reduction.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
\end{center}
\caption{\label{fig:timeComp}
Runtime comparison between intra, inter and inter+inline versions of
our algorithm.
The bars are normalized to the time to run the intra-procedural analysis.
}
\end{figure}

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation is responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our speed, because it increases the
size of the target program.
We have tried our analysis in two different program representations: the
Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
Single Assignment (e-SSA) form~\cite{Bodik00}.
The SSA form gives us a faster, albeit more imprecise, analysis.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/wideningPrec}
\end{center}
\caption{\label{fig:wideningPrec}
The impact of the widening strategy on precision.
Each bar gives precision in \%bitwidth reduction.}
\end{figure}
